|

### 2024-11-25 manuscript/paper.Rmd
addressing Greg's changes and additions. Some questions back to you.

Ecological release theory suggests that a population’s niche should change when important species interactions like resource competition or predation are relaxed or removed (reviewed in @Herrmannetal2021). Removal is posited to create ecological opportunity—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use new resources (@ParentandCrespi2009; @Herrmannetal2021). Ecological release may then be followed by adaptive morphological evolution as traits change to reflect the new niche (@ParentandCrespi2009; @Herrmannetal2021).

For example, a population undergoing ecological release can experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011). This is predicted to result in intersexual divergence in habitat use and associated phenotypes (@Schoener1968; @Shine1989; @BolnickandDoebeli2003; @Butleretal2007; @BolnickandLau2008; @Cooperetal2011; but see @Stuartetal2021; @Blain2022). Such competition-driven “intraspecific character displacement” between the sexes is therefore one explanation for the evolution of sexual dimorphism (@PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleandRowe2017, @DeLisleetal2018).

Much of the theory and empirical data for character displacement between the sexes is based on release from resource competition specifically (e.g., @BolnickandDoebeli2003; @Cooperetal2011; @PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleetal2018). However, release from predation might also drive the evolution of increased sexual dimorphism because an absence of predators should generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).

Here, we tested the prediction that a release from predation results in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (Gasterosteus doryssus). The sequence in this depositional environment is comprised of two lineages (@Bell2009; @Cerasonietal2024). Lineage I was a low-armor form with zero to one dorsal spines and highly reduced pelvises, on average. Lineage I lasted for at least 93,000 years before it was replaced by a second lineage, suddenly, likely on the order of years (@Belletal1985; Stuart et al. unpublished data). Lineage II appeared in the depositional environment fully armored, with complete pelvic girdles, two pelvic spines, and three dorsal spines (@Belletal2006; @Stuartetal2020). It then immediately began evolving adaptive reduction in its armor traits (@Huntetal2008; @Stuartetal2020) until it reached the same low-armor state previously held by lineage I.

The observation of 93,000 years of low-armor stasis in lineage I (@Belletal1985) and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish like trout and other salmonids known to prey on modern threespine stickleback. Armor presence in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly with the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023). In our paleolake basin, only three fossil trout have been found in the same section of rock that has revealed >20,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024).  Thus, we reconstruct an evolutionar history in which it is likely that lineage II migrated from a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024). Lineage II then experienced release from predators in the focal paleolake basin, generating the initial conditions of evolutionary models predicting the evolution of increased sexual dimorphism following ecological release. We test this prediction here.

@@

For our study, we inferred sex in G. doryssus fossils by comparing multivariate morphological trait data from lineage II samples to the same multivariate trait set collected from multiple populations in the closely-related, extant Threespine Stickleback species complex (Gasterosteus aculeatus). Crucially, we determined G. aculeatus sex directly via dissection and/or PCR genotyping. This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive multiple imputation algorithm (@little2002statistical) based on the multivariate morphology of G. aculeatus individuals of known sex. We applied this algorithm to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable. With sex assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework that accounts for uncertainty in sex classification to test for evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.

@@

## Fossil Specimen Data
We used Gasterosteus doryssus data that were previously reported by @Stuartetal2020, @Vojeetal2022, and @Siddiquietal2024. Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Hazen, Nevada, USA. Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years. Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1). Series K started at the previously documented horizon when lineage I was replaced by lineage II. The tempo and mode of lineage II armor reduction during this sequence suggests adaptive evolution by natural selection (@Huntetal2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.

@@

We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol \textcolor{red}{(Whom, personal communication)}.

@@

## Missing data imputation, including fossil sex
Fossil and extant data were combined to create a single data set which contained small amounts on missingness on the observed traits with the exception of sex, which was completely missing for fossil specimens as this was unobservable.  The sex of the fossil data was considered to be missing data and was imputed in the same step of the rest of the missing data using multiple imputation (@little2002statistical).  Imputation was performed using the multiple imputation by chained equations (MICE) algorithm (@MICE) via predictive mean matching (PMM) (@hotdeck) and implemented using the R programming language (@R2022language). Traditionally, the choice for the number of completed data sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis, as was performed here, after imputation.  Therefore, we imputed $M = 100$ complete data sets.  Extant fish records were then dropped from each of the imputed data sets, and only the fossil data was used in downstream analyses of fossil sexual dimorphism.

After performing the Bayesian analysis on each of the $M = 100$ imputed data sets, draws from the posterior distribution based on each of the imputed data sets were pooled together to combine posterior draws across imputations as recommended in @ZhouReiter2010 rather than using Rubin's traditional combining rules.

<!-- Briefly, we used multiple imputation (@little2002statistical) to impute the sex of the fossils. Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.  Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively. We imputed missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis after imputation.  Therefore, we imputed $M = 100$ complete datasets.  -->
\textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
\textcolor{orange}{Updated - GReg}
\textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
\textcolor{orange}{Are we sure we want that here?  I was going to add it in later}

@@

## Modeling fossil stickleback trait means through time for males and females
 The goal of our analysis was to model fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex. We therefore implemented a Bayesian hierarchical model structure (described below) to account for classfication uncertainty and the fact that the rate of change of any one trait mean over time may be a non-constant value.

To estimate fossil sample means through time, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.  In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
\begin{align}
{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

@@

If $Y_{k,ij}$ is a continuous trait, then, potentially conditioning on $Z_{k,ij} = 0$, 
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}
In other words, (\ref{eq:cont}) only considers the subset of fossils where the given trait is present when performing a data analysis. For the traits with potential loss, we model the $Z_{k,ij}$ via a $Bernoulli(p_{k,ij})$ distribution where
\begin{align}
\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
\label{eq:cont_prob}
\end{align}
and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are guaranteed to be between 0 and 1. Modeling traits this way allows us to first estimate the probability of trait loss for an individual at a given time, and ask if the sexes differ in their tendency toward loss through time. Then, for individuals that have not lost the trait, we can estimate trait values and calculate sexual dimorphism through time.

If $Y_{k,ij}$ is a discrete trait, the conventional modelling method is to fit a Poisson distribution. However, the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, so we fitted discrete traits to a generalized Poisson model (@GeneralizedPoisson). Specifically, if $X \sim GP(\lambda,\alpha)$, then 
\begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align}
where GP denotes a generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.

In addition, we modeled two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) to vary with standard length. This is because these are vertebral counts that made along the anterior-posterior access of the fish; these vertebrae contribute to the length of the fish directly and may evolve in a correlated fashion. If $Y_{k,ij}$ is one of the above traits, then
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\lambda_{k,ft_i} = \exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\lambda_{k,mt_i} = \exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

@@

<!-- In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific (potentially log-transformed) trait mean $k$ for females and males, respectively. -->  To account for the possibility that there is a time-dependent (i.e., evolutionary trend) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$, where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, and $u_{k,gt_i}$ is the corresponding residual. For the continuous phenotypes with potential lost trait values, we model $\alpha{k,gt_i}$ in (\ref{eq:cont_prob}) by
\begin{align}
\alpha_{k,gt_i} = \beta_{0,p,kg} + \beta_{1,p,kg}t_i + u_{k,p,gt_i}.
\label{eq:prob}
\end{align}
For convenience, we drop the subscript $p$ for all model descriptions to come, and assume same modelling techniques apply. To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, we defined $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in @OUProcess, the closed form solution for the stochastic differential equation in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant. This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of latent variables.

@@

### Validating Imputation of Sex
```{r eval = FALSE}
library(mixgb)
library(mice)
library(tidyverse)
library(janitor)
library(missMethods)
library(ROCR)

>>>>

Ecological release theory suggests that a population’s niche should change when important species interactions like resource competition or predation are relaxed or removed (reviewed in @Herrmannetal2021). Removal is posited to create ecological opportunity—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use (@ParentandCrespi2009; @Herrmannetal2021). Ecological release may then be followed by adaptive morphological evolution as traits change to reflect the new niche (@ParentandCrespi2009; @Herrmannetal2021).

For example, a population undergoing ecological release can experience novel disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011). This is predicted to result in intersexual divergence in habitat use and associated phenotypes (@Schoener1968; @Shine1989; @BolnickandDoebeli2003; @Butleretal2007; @BolnickandLau2008; @Cooperetal2011; but see @Stuartetal2021; @Blain2022). Such competition-driven “intraspecific character displacement” between the sexes is therefore one explanation for the evolution of sexual dimorphism (@PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleandRowe2017, @DeLisleetal2018).

Much of the theory and empirical data for character displacement between the sexes is based on release from resource competition specifically (e.g., @BolnickandDoebeli2003; @Cooperetal2011; @PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleetal2018). However, release from predation might also drive the evolution of increased sexual dimorphism because an absence of predators should also generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).

Here, we tested this prediction that a release from predation results in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (*Gasterosteus doryssus*). The sequence in this depositional environment is comprised of two lineages (@Bell2009; @Cerasonietal2024). Lineage I was a low-armor form with zero to one dorsal spines and highly reduced pelvises, on average. Lineage I lasted for at least 93,000 years before it was replaced by a second lineage, suddenly, likely on the order of years (@Belletal1985; Stuart et al. unpublished data). Lineage II appeared in the depositional environment fully armored, with complete pelvic girdles, two pelvic spines, and three dorsal spines (@Belletal2006; @Stuartetal2020). It then immediately began evolving adaptive reduction in its armor traits (@Huntetal2008; @Stuartetal2020) until it reached the same low-armor state previously held by lineage I.

The observation of 93,000 years of low-armor stasis in lineage I and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish like trout and other salmonids known to prey on modern threespine stickleback. Armor presence in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly with the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023). In our paleolake basin, only three fossil trout have been found in the same section of rock that has revealed >20,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024).  Thus, we reconstruct an evolutionary history in which it is likely that lineage II migrated from a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024). Lineage II then experienced release from predators in the focal paleolake basin studied here, generating the initial conditions of evolutionary models predicting the evolution of increased sexual dimorphism following ecological release.

@@

For our study, we inferred sex in *G. doryssus* fossils by comparing multivariate morphological trait data from lineage II samples to the same multivariate trait set collected from multiple populations in the closely-related, extant Threespine Stickleback species complex (Gasterosteus aculeatus). Crucially, we determined extant *Gasterosteus aculeatus* sex directly via dissection and/or PCR genotyping. This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive multiple imputation algorithm (@little2002statistical) based on the multivariate morphology of *G. aculeatus* individuals of known sex. We applied this algorithm to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable. With sex assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework that accounts for uncertainty in sex classification to estimate trait means through time. Then we quantified the evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.

@@

## Fossil Specimen Data
We used Gasterosteus doryssus data that were previously reported by @Stuartetal2020, @Vojeetal2022, and @Siddiquietal2024. Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Hazen, Nevada, USA. Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years. Fish from series K were previously measured @Stuartetal2020 for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1). Series K started at the previously documented horizon when lineage I was replaced by lineage II. The tempo and mode of lineage II armor reduction during this sequence suggests adaptive evolution by natural selection (@Huntetal2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.

@@

In 2020, we replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol \textcolor{red}{(Whom, personal communication)}.

@@

## Missing data imputation to classify fossil sex
Fossil and extant data were combined to create a single data set which contained small amounts of missingness on the observed traits \textcolor{red}{Greg, THIS MAY NOT BE TRUE. SOME FOSSIL TRAITS HAD LOTS OF MISSINGNESS. IS THIS A PROBLEM?} with the exception of sex, which was completely missing for fossil specimens as this was unobservable.  WE imputed sex and the rest of the missing data in the same step using multiple imputation (@little2002statistical), performed via the "multiple imputation by chained equations" (MICE) algorithm (@MICE) with predictive mean matching (@hotdeck), and implemented using the R programming language (@R2022language). Traditionally, the choice for the number of completed data sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis, as was performed here, after imputation.  Therefore, we imputed $M = 100$ complete data sets.  Extant fish records were then dropped from each of the imputed data sets, and only the fossil data were used in downstream analyses of fossil sexual dimorphism. 
To test the accuracy of our multiple imputation model in estimating classifying sex,wWe performed 5-fold cross validation using the extant cases only. For a given fold, we took the observed values of sex and removed them \textcolor{red}{Greg, HOW? I GUESS WE REMOVED SEX FOR EVERY INDIVIDUAL?} to reproduce the missingness observed in the fossil data.  Then with one fold held out, we used the remaining folds with observed sex to perform multiple imputation with predictive mean matching, as before, with $M$ = 5,  to build a model that we then used to impute the sex of each fish in the held-out fold (along with any other variables that needed to be imputed).  We then compared the imputed values of sex in each held-out fold to the true values of sex to evaluate the accuracy of the imputation model, and combined the output to estimate the probability of being correctly called male versus female (i.e. the probability of being male was estimated as the proportion of imputations that predicted male \textcolor{red}{Greg, THE PROPORTION OF IMPUTATIONS THAT PREDICTED MALE ***CORRECTLY***?}).  We constructed an ROC curve from these probabilities (Figure \ref{fig:roc}) to visualize imputation accuracy.

\textcolor{red}{THIS STATEMENT ABOUT POOLING DRAWS SEEMS OUT OF PLACE. SHOULDN'T IT GO JUST BEFORE WE TALK ABOUT USING THE POOLED DRAWS?}
After performing the Bayesian analysis on each of the $M = 100$ imputed data sets, draws from the posterior distribution based on each of the imputed data sets were pooled together to combine posterior draws across imputations as recommended in @ZhouReiter2010 rather than using Rubin's traditional combining rules.

<!-- Briefly, we used multiple imputation (@little2002statistical) to impute the sex of the fossils. Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.  Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively. We imputed missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).  Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.  However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis after imputation.  Therefore, we imputed $M = 100$ complete datasets.  -->

@@

## Modeling fossil stickleback trait means through time for males and females
 The goal of our analysis was to model fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex. We therefore implemented a Bayesian hierarchical model structure (described nesxt) to account for classfication uncertainty and the fact that the rate of change of any one trait mean over time may be a non-constant value.

\textcolor{red}We used each for the M = 100 imputed data sets to estimate fossil sample means through time.} For a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.  In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
\textcolor{red}I can't remember. Do we analyze the size-corrected data at all? If not, we should say we left those behind, because we did use those to build the MICE model}
\begin{align}
{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

@@

If $Y_{k,ij}$ is a continuous trait, then, potentially conditioning on $Z_{k,ij} = 0$, 
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}
In other words, (\ref{eq:cont}) only considers the subset of fossils where the given trait is present when performing a data analysis. For the traits with potential loss, we model the $Z_{k,ij}$ via a $Bernoulli(p_{k,ij})$ distribution where
\begin{align}
\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
\label{eq:cont_prob}
\end{align}
and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are between 0 and 1. Modeling traits this way allows us to first estimate the probability of trait loss for an individual at a given time and ask if the sexes differ in their tendency toward loss through time. Then, for individuals that have not lost the trait, we can estimate trait values and calculate sexual dimorphism through time.

If $Y_{k,ij}$ is a discrete trait, the conventional modelling method is to fit a Poisson distribution. However, the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, so we fitted discrete traits to a generalized Poisson model (@GeneralizedPoisson). Specifically, if $X \sim GP(\lambda,\alpha)$, then 
\begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align}
where GP denotes a Generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.

In addition, we modeled two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) to vary with standard length. This is because these are vertebral counts that made along the anterior-posterior access of the fish; these vertebrae contribute to the length of the fish directly and may be correlated with standard length. If $Y_{k,ij}$ is one of the above traits, then
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\lambda_{k,ft_i} = \exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\lambda_{k,mt_i} = \exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

@@

<!-- In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific (potentially log-transformed) trait mean $k$ for females and males, respectively. -->  To account for the possibility that there is a time-dependent (i.e., evolutionary) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$, where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, and $u_{k,gt_i}$ is the corresponding residual. For the continuous phenotypes with potential lost trait values, we model $\alpha{k,gt_i}$ in (\ref{eq:cont_prob}) by
\begin{align}
\alpha_{k,gt_i} = \beta_{0,p,kg} + \beta_{1,p,kg}t_i + u_{k,p,gt_i}.
\label{eq:prob}
\end{align}
For convenience, we drop the subscript $p$ for all model descriptions to come, and assume same modelling techniques apply. To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, we defined $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in @OUProcess, the closed form solution for the stochastic differential equation in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant. This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of latent variables.

@@

### Sex Imputation
```{r eval = FALSE}
library(mixgb)
library(mice)
library(tidyverse)
library(janitor)
library(missMethods)
library(ROCR)
